---
title: "ST558 Project I"
author: "Mana Azizsoltani"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  github_document:
    toc: true
    toc_depth: 1
---

# Required Packages
For this project, I needed to use a variety of really amazing R packages:

  * `knitr`: for creating pretty tables using the `kable()` function.  
  * `tidyverse`: for data manipulation, plot generation (via `ggplot2`), and piping/chaining.  
  * `gridExtra`: to stitch ggplots into a grid.  
  * `rmarkdown`: to knit output files manually using the `render()` function.  
  * `jsonlite`: for pulling data from the various endpoints of the NHL APIs.  

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
# Load necessary packages
library(knitr)
library(tidyverse)
library(rmarkdown)
library(jsonlite)
library(gridExtra)
```

# Creating the Functions  
The first thing that I did was create three functions:  

  1. `getRecords()`: this function takes in an endpoint and team ID (or name) from the NHL Records API and queries the correct data set.  
  2. `getStats()`: this function takes in an optional modifier argument and queries the correct data set from the NHL Stats API.  
  3. `pullData()`: this function is a wrapper function for `getRecords()` and `getStats()`. It serves as a one-stop-shop for the user to access any of the endpoints from above APIs.  

```{r funcs}
# Write URL to access the NHL Records API
getRecords <- function(endpoint, teamID = NULL, teamName = NULL){
  base <- "https://records.nhl.com/site/api"
  table <- endpoint
  if (!is.null(teamName) & is.null(teamID)) {
    full_url <- full_url <- paste0(base, "/", table)
    data <- fromJSON(full_url, flatten = TRUE)$data %>% filter(franchiseName == teamName)
  } else if(!is.null(teamID) & is.null(teamName)) {
      full_url <- paste0(base, "/", table, "?cayenneExp=franchiseId=", teamID)
      data <-fromJSON(full_url, flatten = TRUE)$data
  } else if(is.null(teamID) & is.null(teamName)) {
      full_url <- paste0(base, "/", table)
      data <-fromJSON(full_url, flatten = TRUE)$data
    }
  return(data)
}

# write the function for the NHL Stats API
getStats <- function(modifier=NULL){
  base <- "https://statsapi.web.nhl.com/api/v1/teams"
  if(!is.null(modifier)){
    full_url <- paste0(base, modifier)
    data <- fromJSON(full_url, flatten = TRUE)$teams
  } else {
    data <- fromJSON(base, flatten = TRUE)$teams
  }
  return(data)
}

# Write the wrapper function
pullData <- function(api = c("stats", "records"), endpoint = NULL, 
                     teamName = NULL, teamID = NULL, modifier = NULL){
  if(api == "stats"){
    data <- getStats(modifier = modifier)
    }
  if(api == "records"){
    data <- getRecords(endpoint = endpoint, teamID  = teamID, teamName = teamName)
    }
  return(data)
}

```

# Creating a Nice Data Set  
As a Las Vegas native, I really wanted to do something with Vegas Golden Knights (VGK), so I pulled the 2017/2018 Vegas Golden Knights roster from the `roster` endpoint. Then, since I knew exactly what I wanted in my data, I created two loops to pull two data sets about the players out of the `players` endpoint in the NHL Stats API.  

```{r getData, message=FALSE}
# read in the roster of the VGK for the 2017/2018 season
rosters <- pullData(api = "stats", modifier = "?expand=team.roster&season=20172018")
VGKrost <- as.data.frame(rosters[[30]][[31]])
VGKrost <- VGKrost %>% filter(position.code != "G") %>% select(-person.link)

# Get the stats for each individual player.
ppl <- VGKrost$person.id %>% as.vector(mode = "character")
temp <- data.frame()
df.stats <- data.frame()
for (i in 1:length(ppl)){
  url <- paste0("https://statsapi.web.nhl.com/api/v1/people/", ppl[i],
                "/stats?stats=statsSingleSeason&season=20172018")
  temp <- fromJSON(url, flatten = TRUE)$stats$splits %>% as.data.frame()
  temp$person.id <- as.numeric(ppl[i])
  df.stats <- bind_rows(df.stats, temp)
}

# Get more stats for each individual player.
ppl <- VGKrost$person.id %>% as.vector(mode = "character")
temp <- data.frame()
df.ppl <- data.frame()
for (i in 1:length(ppl)){
  url <- paste0("https://statsapi.web.nhl.com/api/v1/people/", ppl[i])
  temp <- fromJSON(url, flatten = TRUE)$people %>% as.data.frame()
  df.ppl <- bind_rows(df.ppl, temp)
}
```

# Data Manipulation  
To get the data to the form that I wanted, I first had to use an inner join to match-merge the data sets by player ID. The only problem was that there were a bunch of redundant columns or just columns that I knew I didn't need, so I went ahead and removed them. After that, I went ahead and created some new variables.  

```{r manip, message=FALSE}
# Join the three data sets
df.ppl <- df.ppl %>% rename(person.id = id)
dfd <- inner_join(VGKrost, df.ppl, by = "person.id")
df <- inner_join(dfd, df.stats)

# Get rid of redundant rows
df <- df %>% select(-ends_with("link"), -starts_with("primary"), -ends_with("TimeOnIce"), 
                    -position.code, -person.fullName, -firstName, -lastName, -active,
                    -starts_with("birth"), -rosterStatus, -currentTeam.name, 
                    -currentTeam.id, -season, -ends_with("TimeOnIcePerGame"))


# Create new variables
df <- df %>% mutate(domHand = ifelse(shootsCatches=="L", "Left", "Right")) %>% select(-shootsCatches)
df <- df %>% mutate(goalsPerGame = stat.goals/stat.games, 
                    assistsPerGame = stat.assists/stat.games,
                    shotsPerGame = stat.shots/stat.games,
                    hitsPerGame = stat.hits/stat.games) 

```

# Contingency Tables  
I created some tables for the categorical variables that I had in my data set. I was able to see the breakdown of positions, nationalities, and dominant hands on the team.

```{r contTbl}
# Create contingency tables
table(df$position.name) %>% kable(caption = "Number of players in each field position", 
                                  col.names = c("Position", "Frequency"))
table(df$nationality) %>% kable(caption = "Number of Players by Nationality",
                                col.names = c("Nationality", "Frequency"))
table(df$position.name, df$domHand) %>% kable(caption = "Player Position by Dominant Hand",
                                                    col.names = c("Left-Handed", "Right-Handed"))
```

# Quantitative Summaries  
I then created some tables for the mean goals, assists, shots, and hits for the different nationalities and positions. I also created a table that displays the estimated mean and standard error of the amount of goals scored by dominant hand to get an idea to whether or not there was a significant difference.  

```{r quantsums, message=FALSE, warning=FALSE}
# Make the first contingency table: average measurements by position
means1 <- df %>% group_by(position.name) %>% summarize(avgGoals = mean(stat.goals), 
                                                      avgAssists = mean(stat.assists),
                                                      avgShots = mean(stat.shots),
                                                      avgHits = mean(stat.hits))
means1 %>% kable(caption = "Average Season Stats by Position",
                col.names = c("Position", "Goals", "Assists", "Shots", "Hits"))

# Make the second contingency table: average measurements by nationality
means2 <- df %>% filter(nationality %in% c("CAN", "CHE", "USA", "SWE")) %>%
                 group_by(nationality) %>%
                 summarise(avgGoals = mean(stat.goals), 
                           avgAssists = mean(stat.assists),
                           avgShots = mean(stat.shots),
                           avgHits = mean(stat.hits))
means2 %>% kable(caption = "Average Stats for Selected Players Nationalities",
                 col.names = c("Nationality", "Goals", "Assists", "Shots", "Hits"))

# Make the third contingency table: mean & sd by dominant hand
means3 <- df %>% group_by(domHand) %>% summarise(avgGoals = round(mean(stat.goals), 3), 
                                                       sdGoals = round(sd(stat.goals), 3))
means3 %>% kable(caption = "Average Goals (and SE) by Players Handedness",
                 col.names = c("Dominant Hand", "Mean", "Std. Dev."))
```

# Plots
I stitched the four required plots (boxplot, histogram, scatterplot, and barplot) together in a grid. They help visualize different characteristics of the data.  

```{r plotz}
# Create the required plots
plot1 <- ggplot(data = df, aes(x = position.name, fill = domHand)) + geom_bar(position = "dodge") +
           labs(x = "Position Name", y = "Count", title = "Field Positions by Dominant Hand") + 
           scale_fill_manual(values = c("#a8ddb5", "#43a2ca"), name = "Dominant Hand") + 
           theme(axis.text.x = element_text(angle = 45, hjust=1))
plot2 <- ggplot(data = df, aes(x = stat.shots, y = stat.goals, color = domHand)) + 
           geom_point() + scale_color_manual(values = c("#fdb462", "#80b1d3"), name = "Dominant Hand") +
           labs(x = "Season Shots", y = "Season Goals", title = "Shots Taken vs. Goals Scored")
plot3 <- ggplot(data = df, aes(x = position.name, y = stat.assists)) + 
           geom_boxplot(color = "#225ea8") +
           stat_summary(fun = mean, geom="line", group= 1, color= "#a6cee3", size = 1.5) +
           labs(x = "Position Name", y = "Season Assists", title = "Season Assists by Position")
plot4 <- ggplot(data = df, aes(x=hitsPerGame)) + 
           geom_histogram(aes(y = ..density..), binwidth = 1/3, color = "black") +
           labs(x = "Hits Per Game", y = "Density", title = "Distribution of Hits per Game") + 
           stat_density(geom = "line", adjust = .55, lwd = 2, color = "#CC0000")
grid.arrange(plot1, plot2, plot3, plot4, nrow = 2)
```

Then, I created my fifth and final plot to show the the distribution of the goals scored by dominant hand.  

```{r plot5}
plot5 <- ggplot(data = df, aes(x = stat.goals)) + 
  geom_histogram(aes(y = ..density..), bins = 10) +
  stat_density(geom = "line", adjust = 1, lwd = 2, color = "#CC0000") + facet_wrap(~domHand) +
  labs(y = "Density", x = "Season Goals Scored", title = "Distribution of Goals Scored by Dominant Hand")
plot5
```



```{r test}
# players <- fromJSON("https://statsapi.web.nhl.com/api/v1/roster/", flatten = TRUE)
# 
# 
# 
# 
# d <- pullData(api = "records", endpoint = "franchise-goalie-records", teamName = "Boston Bruins")
# 
# d <- getStats("?expand=team.roster&season=20182019")
# d <- getStats("?expand=team.roster&season=20102011")
# d <- getStats("?stats=statsSingleSeason&season=20182019")
# ### TEST STUFF
# 
# full_url <- "https://statsapi.web.nhl.com/api/v1/teams?expand=person.names"
# full_url <- "https://statsapi.web.nhl.com/api/v1/teams?expand=team.stats"
# d <- fromJSON(full_url, flatten = TRUE)$teams
# 
# 
# d <- getStats(modifier = "?expand=team.stats")
# 
# base <- "https://statsapi.web.nhl.com/api/v1/teams"
# d <- fromJSON(full_url, flatten = TRUE)
#   
# base <- "https://records.nhl.com/site/api"
# table <- "franchise"
# attr <- "pikachu/"
# full_url <- paste0(base, "/", table)
# data <-fromJSON(full_url, flatten = TRUE)$data
# 
# table <-"franchise-team-totals"
# full_url <- paste0(base, "/", table)
# data <-fromJSON(full_url, flatten = TRUE)$data
# 
# table <- "franchise-season-records"
# full_url <- paste0(base, "/", table)
# data <-fromJSON(full_url, flatten = TRUE)$data %>% filter(franchiseName=="Vegas Golden Knights")
# 
# table <- "franchise-goalie-records"
# full_url <- paste0(base, "/", table, "?cayenneExp=franchiseId=", id)
# data <- fromJSON(full_url, flatten = TRUE)$data
# 
# table <- "franchise-season-records"
# full_url <- paste0(base, "/", table, "?cayenneExp=franchiseId=", id)
# data <-fromJSON(full_url, flatten = TRUE)$data
```